Towards open collaboration in biomedical and health research: An open collaborative scoping review

### 2025-03-18 doc/paper.qmd
style: :art: small edits, plus reformat Markdown

We also originally intended on searching websites, blogs, and other
online resources, but it was very difficult to programmatically and
systemtically achieve this. So we ended up only searching for scholarly
work that were indexed in databases.

@@

Across the papers found in our literature search, X out of 11 discuss
actual experiences on implementing open collaboration, albeit in
different organizational levels. These range from lab level
[@Turoman2022], center level [@Bush2022; @Network2023], multi-center
level [@ManyPrimates2021; @Network2023] and field level: psychology
[@Alessandroni2022], spinal cord injury [@TorresEspin2021] and
neuroimaging [@Niso2022]. All of these share in common the need for an
infrastructure that promotes open science and open collaboration via
starting building a community of collaborators or making tools (usually
online resources) that researchers can use to reach FAIR (Finndable,
Accessible, Interoperable and Reusable) goals while doing research.
Before delving in those, here we provide the specific needs for
implementing open collaboration in each organizational level, as well as
describing the differences in approaches among experiences between and
across levels (Figure 2).

@@

In Turoman et al the authors describe in detail how the have set up a laboratory whose members collaborate in an open way from study design to dissemination. Here we highlight the role of group meetings as opportunities to propose ideas and have an initial study design proposed. These group meetings are usually started with a section of interesting mistakes, which allows to share the issues, troubleshoot and make the laboratory members comfortable with the fact that errors are unavoidable in science. The feedback from these meetings might lead to updates to a lab manual, which is a great tool for new members to start engaging with the open practices of the lab. This coupled with a support system by senior members of the laboratory allows, in theory, to ease the learning curve of new members to adapt to open practices and collaboration.

In the paper, authors describe their implementation of open practices in each step of the scientific production. Here we focus on the aspects in pre-registration, data analysis and dissemination that improve collaboration between memebers of the labortory and of the same field. For pre-registration they combine slack channels and Trello to communicate between members and OSF for data reposition and uploading pre-registered studies. They put special emphasis on folder structure, which is consistent for all projects, and assigning the authorship in this stage. For data analysis, the authors highlight the figure of the data partner, a member of the laboratory not originally included in the analyses that runs the analyses described in the pre-registered study independently. If the results mismatch with the original ones meetings with the lead analyst or all collaborators are organized to assess what the issue might be. Additionally, this allows for insights that might result in deviations of the original pre-registered study, which should be included and noted in the final manuscript. Importantly, the role of the data partner is described in the final manuscript. Finally, in the dissemination step, they highlight that they might incorporate feedback from presenting their analyses in conferences, which are often included as additional non-preregistered analyses.

Finally, in the paper they also highlight relevant open practices that might not enhance collaboration, but that we consider interesting reviewing too. First, during pre-registration, they allow for papers to contain defined confirmatory and exploratory sections to avoid the usual rigidness of pre-registered studies. Second, in exploratory analyses they included soundness checks whenever possible. Third, the perform sequential hypothesis testing to assess within the limits of a maximum sample size which hypothesis fits the data better without induling in p-hacking. Finally, they collaborate with schools and with universities to disemminate their results.

@@

Interestingly, the level of organization with most papers is the
multi-center level with two papers from ManyPrimates [@ManyPrimates2021;
@Primates2019] and one from the Reproducibility Network [@Network2023].
In this organization level the most urgent needs are: 1) identifying the
needs for open collaboration and 2) creating an infrastructure that
allows those needs to be satisfied. Indeed, the Reproducibility Network
[@Network2023] describes a systematic review as an exploration of the
needs for 11 UKRN centers to implement open collaborative practives. In
this paper they specially focus on tools and resources among the 11
institutions and make clear the need for starting to utilize open
resources as a first step to adapt open collaborative practices. The
first paper from ManyPrimates [@Primates2019] describes the start of
another multi-center collaboration, albeit from different institutions.
It is interesting to observe that in this situation, the complex issue
is to start the collaboration itself, which was ignited thanks to a
symposium and an email chain among future collaborators. In contrast,
ManyPrimates could quickly develop an open collaborative structure with
relative ease in comparison to changing an already solid infrastructure,
like in the case of @Network2023.

>>>>

We also originally intended on searching websites, blogs, and other
online resources, but it was very difficult to programmatically and
systematically achieve this. So we ended up only searching for scholarly
work that were indexed in databases.

@@

Across the papers found in our literature search, X out of 11 discuss
actual experiences on implementing open collaboration, albeit in
different organizational levels. These range from lab level
[@Turoman2022], center level [@Bush2022; @Network2023], multi-center
level [@ManyPrimates2021; @Network2023] and field level: psychology
[@Alessandroni2022], spinal cord injury [@TorresEspin2021] and
neuroimaging [@Niso2022]. All of these share in common the need for an
infrastructure that promotes open science and open collaboration via
starting building a community of collaborators or making tools (usually
online resources) that researchers can use to reach FAIR (Findable,
Accessible, Interoperable and Reusable) goals while doing research.
Before delving in those, here we provide the specific needs for
implementing open collaboration in each organizational level, as well as
describing the differences in approaches among experiences between and
across levels (Figure 2).

@@

In @Turoman2022 the authors describe in detail how the have set up a
laboratory whose members collaborate in an open way from study design to
dissemination. Here we highlight the role of group meetings as
opportunities to propose ideas and have an initial study design
proposed. These group meetings are usually started with a section of
interesting mistakes, which allows to share the issues, troubleshoot and
make the laboratory members comfortable with the fact that errors are
unavoidable in science. The feedback from these meetings might lead to
updates to a lab manual, which is a great tool for new members to start
engaging with the open practices of the lab. This coupled with a support
system by senior members of the laboratory allows, in theory, to ease
the learning curve of new members to adapt to open practices and
collaboration.

In the paper, authors describe their implementation of open practices in
each step of the scientific production. Here we focus on the aspects in
pre-registration, data analysis and dissemination that improve
collaboration between members of the laboratory and of the same field.
For pre-registration they combine slack channels and Trello to
communicate between members and OSF for data deposition and uploading
pre-registered studies. They put special emphasis on folder structure,
which is consistent for all projects, and assigning the authorship in
this stage. For data analysis, the authors highlight the figure of the
data partner, a member of the laboratory not originally included in the
analyses that runs the analyses described in the pre-registered study
independently. If the results mismatch with the original ones meetings
with the lead analyst or all collaborators are organized to assess what
the issue might be. Additionally, this allows for insights that might
result in deviations of the original pre-registered study, which should
be included and noted in the final manuscript. Importantly, the role of
the data partner is described in the final manuscript. Finally, in the
dissemination step, they highlight that they might incorporate feedback
from presenting their analyses in conferences, which are often included
as additional non-preregistered analyses.

Finally, in the paper they also highlight relevant open practices that
might not enhance collaboration, but that we consider interesting
reviewing too. First, during pre-registration, they allow for papers to
contain defined confirmatory and exploratory sections to avoid the usual
rigidness of pre-registered studies. Second, in exploratory analyses
they included soundness checks whenever possible. Third, the perform
sequential hypothesis testing to assess within the limits of a maximum
sample size which hypothesis fits the data better without indulging in
p-hacking. Finally, they collaborate with schools and with universities
to disseminate their results.

@@

Interestingly, the level of organization with most papers is the
multi-center level with two papers from ManyPrimates [@ManyPrimates2021;
@Primates2019] and one from the Reproducibility Network [@Network2023].
In this organization level the most urgent needs are: 1) identifying the
needs for open collaboration and 2) creating an infrastructure that
allows those needs to be satisfied. Indeed, the Reproducibility Network
[@Network2023] describes a systematic review as an exploration of the
needs for 11 UKRN centers to implement open collaborative practices. In
this paper they specially focus on tools and resources among the 11
institutions and make clear the need for starting to utilize open
resources as a first step to adapt open collaborative practices. The
first paper from ManyPrimates [@Primates2019] describes the start of
another multi-center collaboration, albeit from different institutions.
It is interesting to observe that in this situation, the complex issue
is to start the collaboration itself, which was ignited thanks to a
symposium and an email chain among future collaborators. In contrast,
ManyPrimates could quickly develop an open collaborative structure with
relative ease in comparison to changing an already solid infrastructure,
like in the case of @Network2023.
